---
title: "Nested random effects"
author: "Miles"
date: "2025-06-17"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(lme4)
```

https://cambiotraining.github.io/stats-mixed-effects-models/materials/08-nested-random-effects.html

## 8  Nested random effects
Each doctor appears under only one hospital, so it's nested
```{r load health data}
health = read.csv('~/Documents/MDA_GLM/GLM_content/health.csv')
head(health)
head(table(health$doctor, health$hospital), 10)
```
## 8.2.1 Random intercept
$$
outcome_{ijk} = (\beta_0 + b_{0i} + b_{0ij}) + \beta_1 \cdot treat_{ijk} + \epsilon_{ijk} 
$$
where i indicates hosp, j indicates doc, and k patient.

```{r random intercept}
# lme_intercepts = lmer(outcome ~ treatment + (1|hospital) + (1|hospital:doctor), data=health)  # correct
lme_intercepts = lmer(outcome ~ treatment + (1|hospital) + (1|doctor), data=health)
summary(lme_intercepts)
```
Correct syntax should be the first one, (1|hospital:doctor),  but the webpage uses the second one, (1|doctor), which is also valid if the data structure clearly shows that doctors are uniquely associated with one hospital (i.e., no doctor appears in multiple hospitals).

## 8.2.2 Random slope
What level to fit random slopes at depends on the what level the treatment/predictor varies at.

### Predictor/treatment varies at level 1 (between patients, within doctors)
The model is 
$$
outcome_{ijk} = (\beta_0 + b_{0i} + b_{0ij}) + (\beta_1 + b_{1i}  + b_{1ij}) \cdot treat_{ijk} + \epsilon_{ijk} 
$$
```{r random slope}
lme_slopes = lmer(outcome ~ treatment + (1+ treatment | hospital) + (1+treatment | hospital: doctor), data=health)
# (1+treatment | hospital: doctor) == (1+treatment | doctor) 
# both give the same result
summary(lme_slopes)
```

### Predictor/treatment varies at level 2 (between doctors, within hospitals)
```{r load health2 data}
health2 = read.csv('~/Documents/MDA_GLM/GLM_content/health2.csv')
head(health2, 10)
head(table(health2$doctor, health2$hospital), 10)
```

Now imagine scenario where each doctor is a specialist in a certain type of treatment (surgery or drug), but cannot deliver both. So treatment does not vary within doctor but only varies within hospital. The model now should be  
$$
outcome_{ijk} = (\beta_0 + b_{0i} + b_{0ij}) + (\beta_1 + b_{1i}) \cdot treat_{ijk} + \epsilon_{ijk} 
$$
$b_{1ij}$ measures the effect of treatment on outcome attributed to doctor $j$. There is no $b_{1ij}$ before $treat_{ijk}$ b/c each doctor only uses one treatment.
 
```{r new model}
lme_slopes = lmer(outcome ~ treatment + (1+ treatment | hospital) + (1 | hospital: doctor), data=health)
```

### Predictor/treatment varies at level 3 (between hospitals)
Now imagine scenario where each hospital can offer only one treatment. So treatment does not vary within the hospital but between hospitals. In this case there is no $b_{1i}$ before $treat_{ijk}$, b/c $b_{1i}$ measures the effect of treatment on outcome arising from hospital $i$.

$$
outcome_{ijk} = (\beta_0 + b_{0i} + b_{0ij}) + \beta_1 \cdot treat_{ijk} + \epsilon_{ijk} 
$$
This goes back to random intercepts only model.

## 8.3 Implicit vs. explicit nesting

```{r Paste data}
data('Pastes')
head(Pastes)
```
Response variable is strength. There's no fixed effects. Clustering variables are batch/delivery (i= A, B, ..., J), and cask (j=1,2,...,30). Cask is *nested* with batch. From each sample collected per cask, there are 2 assays (k=1,2)

Model the strength, adjusting for variation across batches and casks.
$$
strength_{ijk} = \beta_0 + b_{0i} + b_{0ij} + \epsilon_{ijk}
$$
This is a nested design and we actually have 30 unique casks (3 per batch). (Coding doesn't determine the design, experimental reality does.) 

Therefore, use `(1|batch:cask)` instead of `(1|bcask)`. Can also use A/B syntax `(1|batch/cask) == (1|batch) + (1|batch:cask)`
```{r fit}
lme_paste_wrong = lmer(strength ~ 1 + (1|batch) + (1|cask), data=Pastes) # wrong
summary(lme_paste_wrong)

lme_paste_explicit = lmer(strength ~ 1 + (1|batch) + (1|batch:cask), data=Pastes)  # correct, explicit nesting
summary(lme_paste_explicit)
```
Number of obs: 60, groups:  batch, 10; cask, 3
Number of obs: 60, groups:  batch:cask, 30; batch, 10

The latter correctly identifies the number of casks. Always check # groups per clustering variables in the output.


## Exercise: cake dataset
```{r data}
data("cake")
head(cake,10)
table(cake$replicate, cake$recipe)
```
Response is angle. Fixed effects of interest are recipe (i=1,2,3) and temperature (k=1,2,...,6). Clustering variable is replicate (given i, j=1,2,..,15 OR j=1,2,...,45). Which is completely nested within the recipe, despite the table. 

$$
angle_{ijk} = (\beta_0 + b_{0ij}) + \beta_1 \cdot recipe_{i} + \beta_2 \cdot temp_{ijk} + \beta_3 \cdot recipe_{i} \cdot temp_{ijk} + \epsilon_{ijk}
$$

```{r model 1: random intercept only}
lme_cake1 = lmer(angle ~ recipe * temperature + (1|recipe:replicate), data=cake)
summary(lme_cake1)
```

$$
angle_{ijk} = (\beta_0 + b_{0ij}) + \beta_1 \cdot recipe_{i} + (\beta_2 + b_{1ij}) \cdot temp_{ijk} + \beta_3 \cdot recipe_{i} \cdot temp_{ijk} + \epsilon_{ijk}
$$
```{r model 2: try adding random slope; temperature as a factor}
lmer(angle ~ recipe * temperature + (1 + temperature|recipe:replicate), data=cake)
```
Error is "number of random effects (=270) for term (1 + temperature | recipe:replicate)" 
Here temperature is a factor with 6 levels so it is 5 dummies. The model actually is `(1 + temperature175 + ... + temperature215 | recipe:replicate)`, which is 6 random effects * 45 batches (j=1,2,...,45) = 270

The solution is use numeric temperature:
```{r model 3: add random slope by using numeric temp}
lme_cake3 = lmer(angle ~ recipe * temp + (1 + temp|recipe:replicate), data=cake)
summary(lme_cake3)
```

Q: Why might we prefer to use the numerical/continuous version?
A: 
1. Few params. using temperature as a factor creates 5 dummy variables and thus 5 slopes, but when using as a continuous variable there's only 1 slope. It also enables random slopes (see model 2 and 3), 

2. Better for interpretation and interpolation/prediction.

3. If the realtionship between temperature and outcome is nonlinear or irregualr across the levels, don't use the numeric version.





